### 2017-06-25 Manuscript/Response_to_reviewers.Rmd

**June 2017**

\bigskip

Dear Ben,

Thank you for taking the time to consider our manuscript for publication in *Journal of Animal Ecology*. We were very pleased that the editors and reviewers were all so enthusiastic about the quality of our paper, and were impressed by the detailed and constructive nature of the reviews. We have now been through the Associate Editor and reviewer comments in detail, and made substantial changes to our manuscript in response. Below we outline, point-by-point, how we have responded to each of the AE and reviewer comments. We also reproduce the sections of our manuscript we have changed in response to the comments. AE/Reviewer comments are in bold, and reproduced manuscript sections are indented and italicised.

@@

\bigskip

**This manuscript describes spatio-temporal variation in telomere dynamics in the cooperatively breeding Seychelles warbler. It uses an impressive long-term longitudinal dataset of telomere lengths (obtained by qPCR) of >400 individuals and relates telomere length and dynamics to age, sex and various ecological variables, such as yearly insect availability and density, local territory quality and group size and the number of helpers during reproduction.**

**The manuscript has now been read by three reviewers and myself. All of us are very impressed with the dataset and questions asked and answered in this manuscript. Two of the reviewers, however, have some important methodological questions, and some suggestions for updating the bibliography of the manuscript. I agree with their assessment, and believe that when the authors thoroughly clarify these issues, their paper will make an important contribution to the field and will be widely read and cited.**

RESPONSE: We thank the editor and for these comments. We have now addressed the methodological questions raised by reviewers one and three, and have updated several of our analyses in line with their suggestions. In particular, we have now included a qPCR plate ID random effect term into all relevant analyses, which has not changed our results. We have also performed additional analyses to show that sample storage time does not affect telomere length, included more information on the DNA quality of our samples, including a new supplementary figure. We have also updated our discussion to address the methodological concerns of reviewers one and three, and we have toned down our conclusions on one of the specific analyses, as reqwuested by reviewer one. Finally, we have updated our bibliography, in accordance with the reviewer comments. We believe that incorporating the reviewer comments has substantially improved the clarity, readability and rigour of our work.

@@

---

\bigskip

**You should mention in the methods how the samples were assigned to the different plates (i.e. totally random, balanced between years/age, or all samples of 1 individual on the same plate?). If you allocated more or less randomly your samples to the different plates, can you check (and present as supplementary) that you do not have a “plate effect” on the average TL measurement. If you do observe such plate effect (something not that unusual with qPCR...), it might contribute to explain your low repeatability, and maybe part of the telomere lengthening that you describe.**

RESPONSE: We are confident that plate effects are not a source of systematic bias in our data. Firstly, we assigned samples completely randomly (i.e. using a random number generator) to plates prior to running the telomere assays, minimising the chance that plate effects would introduce systematic differences among ages, sexes or cohorts. Secondly, our method of estimating RTL normalises samples across plates, as each RTL value is obtained by subtracting sample Cq values from Cq values obtained from a "calibrator" sample, which is run on every plate (information contained within Bebbington et al. 2016, cited). However, we appreciate that plate effects could still occur (e.g. because of an effect on the calibrator sample), so we have now included plate ID as a random effect in our models. Our results are almost identical, with all the reported significant effects remaining. The only change is that the cohort x age effect on telomere length is now stronger (see response to later comment). Overall, we can be very confident that plate effects have not biased our results. We have updated the methods sections in the relevant places to describe the addition of the plate ID random effect.

@@

With regards to the observed telomere lengthening, it is very possible that inter-plate variability accounts for some of the observed lengthening in our dataset, but it cannot account for the amount and extent of lengthening we find. Indeed, this is exactly what Figure 2B shows. The lengthening observed "within" samples can be attributed to error across plates, but the additional lengthening observed "among" samples cannot be explained by measurement error.

We now explain our randomisation approach in the methods, lines 168-170:

@@

\begin{adjustwidth}{2cm}{2cm}

\textit{
"Prior to qPCR, we used a random number generator to assign samples to qPCR plates, to ensure that no systematic bias could occur with regards to age, sex, cohort or ecological environment."
}

\end{adjustwidth}

We also include a discussion on error rate on lines 413-423:

@@

\begin{adjustwidth}{2cm}{2cm}

\textit{
"Measurement error is one factor that is an issue in our study, and other studies that use qPCR to measure telomere length (Nussey et al. 2014). The measurement error in our study has most likely arisen because samples were run aver a period of several years, during which reagents, consumables and personnel all change. Compared to a set-up where a small amount of samples are all run at the same time, it is unsurprising that we have a higher error rate. Techniques for measuring telomere length with a greater degree of precision may prove helpful in future ecological studies of telomere dynamics, but at present there is still a trade-off between obtaining precise telomere measurements, and utilising the large sample sizes necessary for ecological study. Further, measurement error is unlikely to be a cause of type I error in our study, because we were careful to randomise all samples across qPCR plates, and normalise RTL estimates across plates. However, the noise associated with within and among plate measure may have resulted in a decrease in explanatory power, and possibly in a degree of type II error."
}

@@

---

\bigskip

**1.78 (78%) would not be considered appropriate by many researchers for qPCR methodology (the commonly accepted limit is more or less + or minus 10%, so here between 90% (1.90) and 110% (2.10)). Your efficiency is more than 20% below the expected 100%, which raises some questions about the validity of your methodology. How can you be confident that this low efficiency is not affecting the quality of your data?**

RESPONSE: Again, the level of efficiency is lower than we would have liked, although this value sits fairly squarely within that found in the published telomere literature (see Fairlie et al. 2015, for a recent example), and often it is not even reported. We also note that this problem is not expected to be too drastic, because variation in efficiency is taken into account when calculating RTL from the Cq values. And again, we suggest that because of our study design, random error of this kind makes the majority of our analyses conservative, and is accounted for in the analysis of telomere lengthening.

@@

---

\bigskip

**Important information is missing in the methods section regarding the age and life stage at which birds have been sampled. Telomeres might shorten in a few days during the nestling period (e.g. Stier et al. 2015b, 2016), and might be affected by breeding effort (e.g. Reichert et al. 2014, Sudyka et al. 2014), so it is important in my opinion to know more about the age and status of the birds used in this study.**

RESPONSE: We now provide more clarification about how we determined and analyse age, lines 128-133:

@@

\begin{adjustwidth}{2cm}{2cm}

\textit{
"The majority (96\%) of individuals are ringed (with an individually numbered metal ring and unique combination of colour rings) within the first year of life, and so are of known age. We aged all birds using information on eye colour at first capture (Komdeur et al. 1991) and previous capture history (Richardson et al. 2003). Within the first year of life, birds are classified as nestlings less than one month old (rounded to one month for analyses), fledglings less than six months old (rounded to six months) or subadults up to one year old (rounded to 10 months). Ages for adult birds were rounded to the nearest year."
}

@@

\begin{adjustwidth}{2cm}{2cm}

\textit{
We found no evidence of DNA degradation in older samples (Fig. S1). 
}

\end{adjustwidth}

We also performed an analysis of RTL and sample storage time, using adult birds, and found no evidence of a relationship - lines 254-255.

@@

**Cohort effects (year of birth in analyses of nestling TL) have been reported without being clearly discussed (e.g. Stier et al. 2014, Becker et al. 2015). It might be worth mentioning it in the discussion, and pointing out the difficulties in distinguishing between real cohort effects and potential methodological artefacts linked to sampling, storage, extraction and/or qPCR methods. Were the different cohorts well randomized between qPCR plates?**

RESPONSE: See response above re. randomisation. While not an issue for our study, we agree that it is a point worth raising. We now discuss this issue on lines 376-383:

\begin{adjustwidth}{2cm}{2cm}

\textit{"Although a few studies have shown that temporal variation in telomere dynamics occurs in natural populations, these have been limited in the number of seasons they cover (Mizutani et al. 2013; Watson, Bolton, and Monaghan 2015; Fairlie et al. 2016). Other studies have found cohort effects but not discussed them in an ecological context (Stier et al. 2014; Becker et al. 2015). One problem with studying cohort effects is that it can be difficult to tease apart true cohort effects from effects that may arise if samples degrade with storage time, and/or batch effects in telomere assays, although neither of these factors were a problem in our study. Indeed, the long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods."}

@@

---

\bigskip

**You mention in several places that telomere dynamics with age is different between cohorts (e.g. lines 241-244; 299-300; 350-54), but in my opinion this is not supported by your data; so such conclusion should be avoided. Indeed, in the first analysis of RTL, including the log age * cohort interaction only marginally improved the fit of the model (delta AIC of 1.40, lines 242-244), which is not sufficient to support your conclusion. In addition, you mention that including cohort ID in the within-individual analyse of TL was even decreasing the fit of the model to your data, again suggesting that telomere dynamics is not significantly impacted by the cohort. Please remove any reference to such effect or try to provide convincing analysis justifying your conclusion.**

RESPONSE: Interestingly, this is the one effect that has changed from  including a plate ID random effect in our analyses, with delta AIC now ~11. However, we do not wish to place any strong conclusions on this analysis, given that the significance of the effect appears to be dependent on the inlcusion of a plate ID random effect that we had not previously considered. Thus, we agree that while there is convincing evidence for cohort effects on telomere length, the evidence for cohort effects on telomere dynamics is weaker. We have therefore removed references to this effect throughout the discussion, and allow readers to interpret the effect as they wish.

@@

---

\bigskip

**While I agree that your results are very interesting and biologically meaningful (age effect, sex effet, link with body size and food abundance), and that they are very similar to those on Soay sheep from Dan Nussey’s team, an intra-individual repeatability of 8% is a bit worrying in my opinion. This is especially worrying for using telomere length/dynamics as a “biomarker” as you suggest it in several places in the manuscript. A true biomarker should be at least to some extent consistent within an individual to be useful. You should raise such limitation in your discussion.**

RESPONSE: We are not sure that "worrying" is the correct phrase to use, as we anticipate that other studies in natural population will find similar patterns. Further, we disagree with the sentiment that a biomarker must be consistent over individual lifespans to be useful - rather, a useful marker needs to change with ecological context. We feel that we cover this point in the discussion, lines 366-374:

@@

\begin{adjustwidth}{2cm}{2cm}

\textit{"Such inconsistent changes in telomere length over lifespans could occur due to changes in the cellular composition of of the blood within individual samples, or due to the actual elongation of telomeres (Blackburn et al. 1989). Determining the mechanism of these changes is essential for how we view telomeres as biomarkers of costs. For example, if telomeres can be lengthened in response to improvements in environmental conditions, this would suggest that they reflect short- to medium-term costs, rather than the cumulative costs that an individual has faced over its lifespan (Bateson 2016). New statistical and technical approaches are therefore now required to determine the mechanisms behind increases in telomere length within individuals, so that biologically informed hypotheses about the ecological causes and consequences of these increases can be generated and tested."}

@@

---

\bigskip

**You mention that in other bird studies such within-individual consistency was much higher because they were done in the lab/nest-boxes. Another major difference with these studies is the shorter time-frame on which they were conducted, in addition to the more controlled conditions to store the samples. Therefore you would expect less noise in theses studies linked to methodological artefact due to sample storage/integrity, which might contribute to explain your low within-individual consistency.**

RESPONSE: It is true that the shorter timescales of these studies may be the reason for the higher within-individual consistency, and we have now added a note to this effect, lines 348-359:

@@

\begin{adjustwidth}{2cm}{2cm}

\textit{"Consistent with a pattern of sporadic changes in telomere length with age, we found that within-individual telomere measurements were only weakly correlated. Although some of this low within-individual repeatability will occur due to measurement error, our within-sample repeatability was still much higher than our within- individual repeatability. Such a low value of within-individual repeatability in telomere length is in contrast to other avian studies in which within-individual telomere length measurements were highly consistent, and individual-level telomere shortening occurred throughout the juvenile period and into adulthood (Heidinger et al. 2012; Boonekamp et al. 2014). However, the lifelong telomere dynamics found in Seychelles warblers are strikingly similar to those found in Soay sheep (Fairlie et al. 2016). This discrepancy in results may be because in our study, and that of Fairlie et al. (2016), individuals were born and reared in the wild, as opposed to in nestbox or laboratory conditions. Alternatively it may be because our longitudinal telomere measurements have been taken over longer time periods."}

@@

\end{adjustwidth}

However, as we argue above, we do not think it is likely that the pattern is a result of sample storage/DNA integrity. We have no evidence from our samples that there is sample degradiation with respect to DNA integrity or telomere length (see above).

@@

**Line 16: Telomeres shorten in response to oxidative stress and cellular division. It is important to mention that here since you find higher telomere shortening very early in life that might be linked to cellular division rather than oxidative stress.**

RESPONSE: Yes, this is a good point. We now clarify in the discussion, lines 328-332:

\begin{adjustwidth}{2cm}{2cm}

\textit{"Our study adds to the substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life - most likely as a consequence of the much more rapid rate of cellular division that occurs during the growth phase, but also perhaps higher levels of cellular stress during development (e.g. Frenck, Blackburn, and Shannon 1998; Haussmann, Vleck, and Nisbet 2003; Heidinger et al. 2012)."}

@@

---

\bigskip

**Lines 55-56: While I acknowledged that Von Zglinicki 2002 is the common citation used by people to support the role of oxidative stress in telomere shortening, this is in vitro work with cultured cells using supra-physiological doses of ROS. I would suggest to mention that this relationship is observed “in vitro” or provide a reference supporting such assumption in vivo (if you find one, because I am still looking for solid evidence of this relationship..).**

RESPONSE: Agreed. We now have clarified that this is in vitro.

@@

---

\bigskip

**Lines 231-234: information on sample size, efficiencies and repeatability would be better placed in the methods section in my opinion, since it is not results per se.**

RESPONSE: We tried moving this section as suggested, but could not do so without significantly disrupting the flow and readability of the manuscript. We have therefore left it as it was, but can change if this is something viewed as essential.

@@

---

\bigskip

**Lines 267-271: You mention that “longitudinal telomere length increased with log age...”, and cite fig 2C, 2D to support your statement. As far as I understand fig 2C, it seems that telomere length decrease with age before more or less 2 years of age, and then is not changing overall with age later in life. Yet, such figure does not fit with the estimate you give in the text, so the presentation of your results is quite confusing here.. I do not see how your results are supporting this particular statement. In my opinion, you should probably state that the probability of detecting a within-individual telomere shortening is decreasing with age // the probability of detecting telomere elongation is increasing with age.**

RESPONSE: We agree that this was confusing, and have simplified along the suggested lines - now lines 291-295:

@@

\begin{adjustwidth}{2cm}{2cm}

\textit{"We found that the likelihood of telomere lengthening increased with log age (GLMM with lengthened yes/no as binomial response; estimate = 0.296, CIs = 0.005, 0.588). Increases in telomere length were most likely to be observed shortly after the juvenile period, at around four years of age, and later in life (although sample sizes for older birds are much smaller; Fig. 2C,D)."}

@@

---

\bigskip

**Line 302: the sentence is too vague, especially “particular life stage”.**

RESPONSE: We have now simplified the sentence to make it clearer, lines 323-325:

@@

**Line 321: same comment as above “specific periods”.**

RESPONSE: We are not completely sure what is unclear about this, but have tried rephrasing. Now lines 346-347:

\begin{adjustwidth}{2cm}{2cm}

\textit{"Increases in telomere length were not consistent over individual lifespans, but occurred in bouts, against a backdrop of overall lifelong telomere shortening."}

@@

**You have an impressive dataset of longitudinal data and your work on telomere dynamics can be an important contribution to the current knowledge. However, I have some major concerns regarding your manuscript.**

**Your introduction could be improved, first of all, by citing correctly. Also, there is some more recent literature on the topic, which you could cite. See more in comments on the introduction below. Further, I have some methodological questions. I would like to see more information on how samples were distributed over qPCR plates. Were samples of the same individual on the same plate? If not, how did you take plate ID into account in your statistical analyses? How exactly did you relate territory quality to RTL and delta RTL change? Only for birth year? Reading your abstract and aims in the introduction, I expected more. In the discussion, you speculate on effects of environmental conditions on telomere dynamics. Is there a relationship with telomere lengthening between years and insect abundance/territory quality in the year of lengthening? I guess you could test this with the data you have? This could add to your discussion on potential telomere lengthening, which remained rather methodological. In my opinion, the ecological context in the discussion could be improved. More comments see below.**

**I apologise for a delayed submission of my review, but hope that you find my comments helpful!**

RESPONSE: We are pleased that the reviewer is generally positive about our manuscript. We thank the reviewer for pointing out areas where our citation of previous work was incomplete - although we believe our original citations were generally correct, we agree that there were one or two mistakes, and that it is important to provide the reader with a complete context for our work. We respond this point, and to each of the individual questions/comments, below.

@@

**L 65-66: Here you state there is little evidence of a direct relationship. Only a few lines above, in sentence line 59-61, you write about a potential causal relationship. Please consider to rewrite this, so that it is in one paragraph.**

RESPONSE: We have re-written the text so that the contrast is clearer here. Lines 60-71:

\begin{adjustwidth}{2cm}{2cm}

\textit{"When telomeres become critically short, cells senesce (Campisi 2003), and the accumulation of these cells has been suggested to result in organismal senescence and death (Wong et al. 2003). The association between telomere length and senescence has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival or lifespan have been documented in wild populations of several species (Barrett et al. 2013; Stier et al. 2015). As yet, however, there is little direct evidence that the relationship between telomere dynamics and survival is causal (Simons 2015).}

\textit{"Although the causal role of telomeres in senescence and survival is not yet clear, there is mounting evidence that telomeres can act as biomarkers of individual condition and ageing in wild populations. Specifically, telomeres may be able to provide a measure of the ecological stress that an individual has experienced - a signature that can otherwise be difficult to detect (e.g. Schultner et al. 2014; Asghar et al. 2015; Bebbington et al. 2016)"}

@@

---

\bigskip

**L136: You describe how you assess territory quality etc. But how do you relate it to RTL or delta RTL in your statistical analyses? Do you only use territory quality in the hatching year? You could relate territory quality between RTL measurements to telomere dynamics. Is the variation of territory quality within-individually over time smaller than among-individually? Is the difference between years much larger than between territories?**

### 2017-06-25 Manuscript/SW_early_life_telomeres_FULL.Rmd
Edits

##Abstract
1. Understanding individual-level variation in response to the environment is fundamental to understanding life-history evolution and population dynamics. Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival and lifespan. Investigating telomere dynamics may help us quantify individual variation in the costs experienced from social and ecological environmental factors, and enhance our understanding of the dynamics of natural populations.

@@

Telomeres have been proposed to be a potential biomarker of such costs [@Monaghan2014]. Telomeres are repetitive DNA sequences at the ends of linear chromosomes that protect against DNA damage. Telomeres generally shorten with age [@Muezzinler2013; @Barrett2013a], and there is evidence from a range of taxa that telomere shortening is fastest in early life [e.g. @Frenck1998; @Heidinger2012]. *In vitro* research has shown that telomere shortening can be accelerated by oxidative stress  [@VonZglinicki2002], which can be elevated due to many environmental factors. There is evidence from humans, and from captive and wild animal populations, that telomere shortening is influenced by the conditions experienced during both early life and adulthood  [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015]. Importantly, the extent of telomere shortening is linked to senescence and survival. When telomeres become critically short, cells senesce [@Campisi2003], and the accumulation of these cells has been suggested to result in organismal senescence and death [@Wong2003]. The association between telomere length and senescence has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival or lifespan have been documented in wild populations of several species [@Haussmann2010; @Barrett2013a; @Stier2015]. As yet, however, there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015].

Although the causal role of telomeres in senescence and survival is not yet clear, there is mounting evidence that telomeres can act as biomarkers of individual condition and ageing in wild populations, providing a measure of the ecological stress that an individual has experienced, a signature that can otherwise be difficult to detect [e.g. @Schultner2014; @Asghar2015; @Bebbington2016]. There is also evidence that telomere length, measured longitudinally in individuals, can increase as well as decrease [@Simons2014; @Bateson2016a], which has important ramifications for our understanding of how telomeres reflect costs. However, such increases in telomere length are often attributed to measurement error [@Steenstrup2013; but see @Bateson2016a], and as such their ecological significance is unknown.

@@

In this study, we test how lifelong telomere dynamics are related to environmental variation across 22 Seychelles warbler cohorts (years). We first study how telomere length and rates of shortening are related to age and sex across all life stages, and how this relationship varies among cohorts, in order to gain an in-depth understanding of the temporal dynamics of telomere changes. We then examine, within individuals, how telomere length changes with age, and statistically test whether observed increases in telomere length across temporally longitudinal samples within individuals are larger than can be accounted for by measurement error. Finally, we test how telomere length and shortening are related to a wide range of social and environmental variables in order to gain a fuller understanding of the forces driving telomere dynamics in natural populations.

@@

The Seychelles warbler is a small (~15 g), insectivorous passerine bird with a mean life expectancy of 5.5 years at fledging [@Hammers2013]. The population of *ca* 320 adult birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Komdeur1992; @Richardson2003b; @Spurgin2014a]. This species' main breeding season runs from June-September (though a small proportion of pairs also breed between January–March), when the breeding females on many of the *ca* 110 territories will attempt to breed, laying one or, rarely, two or three eggs [@Komdeur1991]. Breeding attempts are often unsuccessful, and as a result of this low reproductive output, and higher mortality in first-year birds [39% in first-year birds versus 16% in adults; @Brouwer2006], cohort sizes in the Seychelles warbler are typically small (< 50). The 22 hatch year cohorts used in this study cover `r min(dd$LayYear)` to `r max(dd$LayYear)` – the time period during which our data and sampling are most complete.

The majority (96%) of individuals are ringed (with an individually numbered metal ring and unique combination of colour rings) within the first year of life, and so are of known age. We aged all birds using information on eye colour at first capture [@Komdeur1991a] and previous capture history [@Richardson2003b]. Within the first year of life, birds are classified as nestlings less than one month old (rounded to one month for analyses), fledglings less than six month old (rounded to six months) or subadults up to one year old (rounded to 10 months). Ages for adult birds were rounded to the nearest year. As Seychelles warblers are non-migratory endemics naturally confined to the island [@Komdeur2004a], an extensive biannual census of birds on Cousin during each breeding season gives accurate measures of local density, social status (e.g. breeder, helper, non-helper) and individual survival [@VandeCrommenacker2011; @Barrett2013a]. Full details of monitoring methods can be found in Brouwer *et al.* [-@Brouwer2012].

@@

For a large subset of birds we had longitudinal data, with two or more samples taken at different ages (n = `r length(unique(c(dd3$BloodID,dd3$BloodID1)))` measurements from `r length(unique(dd3$BirdID))` birds). For these individuals we calculated the within-individual change in RTL by subtracting RTL at time point *t* from RTL at time point *t* + 1 (hereafter $\Delta$RTL, n = `r nrow(dd3)` measurements). Negative values of RTL reflect decreases in telomere length with age, while positive values reflect increases. Individuals were not always caught at the same time of year, but within a 3 month breeding season window.

@@

Using the longitudinal data, we then tested how telomeres change with age in individuals, using LMMs of RTL as a response and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009] as an explanatory variable. We calculated $\Delta$age using log and polynomial transformed age data, and carried out model selection as above, with the exception that we did not model $\Delta$age as a factor (due to a lack of discreet groupings), and mean age was also included in models to partition within-individual *vs* cross-sectional effects [@VandePol2009].

@@

When examining the distribution of longitudinal telomere changes we observed some increases in telomere length with age in individuals. We therefore repeated the qPCR on each sample  using completely separate reactions run on separate plates and used these repeat measurements to test whether these increases could be explained by measurement error. We calculated the change in RTL between pairs of repeat measurements within the same samples (hereafter $\Delta$RTL~*sample*~; N = `r nrow(subset(ddL,Group == 'Within samples'))` pairs of measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` birds) in exactly the same way as for across samples (hereafter $\Delta$RTL~*individual*~). To test whether greater changes in RTL were observed among individuals compared to among repeat samples, we compared the variance in $\Delta$RTL~*sample*~ and $\Delta$RTL~*individual*~ using a Levene's test. Then, to separately test whether the extent of telomere increases and decreases within individuals were greater than expected by measurement error, we split $\Delta$RTL measurements into groups in which RTL decreased ($\Delta$RTL < 0) and increased $\Delta$RTL ($\Delta$RTL > 0), and tested whether $\Delta$RTL~*individual*~ values were significantly different from $\Delta$RTL~*sample*~ values, using Wilcoxon tests.

We also tested whether consistent telomere lengthening across our dataset using a modified version of the approach developed by Simons *et al.* [-@Simons2014]. Briefly, this approach utilises samples with at least three telomere measurements to compare residual variance in telomere change over time with the overall change in telomere length between the first and last telomere measurements [@Simons2014]. If, in samples that increase in length, the overall increase in telomere length exceeds the residual variance, then telomere lengthening cannot be explained by error [@Simons2014]. If, on the other hand, increases in telomere length are due to measurement error, within-individual residual variance in telomere length is expected to be similar to overall observed increases in telomere length.

@@

For individuals with longitudinal data we repeated the above analyses of telomere dynamics, replacing telomere length with $\Delta$RTL~*individual*~ as the response variable, and including the environmental/social explanatory variables from the first of the two sampling points. We excluded the cohort ID random effect from this analysis, as longitudinal telomere dynamics did not differ among cohorts; see results.

>>>>

##Abstract
1. Understanding individual-level variation in response to the environment is fundamental to understanding life-history evolution and population dynamics. Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival and lifespan. Investigating telomere dynamics may help us quantify individual variation in the costs experienced from social and ecological factors, and enhance our understanding of the dynamics of natural populations.

@@

Telomeres have been proposed to be a potential biomarker of such costs [@Monaghan2014]. Telomeres are repetitive DNA sequences at the ends of linear chromosomes that protect against DNA damage. Telomeres generally shorten with age [@Muezzinler2013; @Barrett2013a], and there is evidence from a range of taxa that telomere shortening is fastest in early life [e.g. @Frenck1998; @Heidinger2012]. *In vitro* research has shown that telomere shortening can be accelerated by oxidative stress  [@VonZglinicki2002], which can be elevated due to many environmental factors. There is evidence from humans, and from captive and wild animal populations, that telomere shortening is influenced by the conditions experienced during both early life and adulthood  [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015]. Importantly, the extent of telomere shortening is linked to senescence and survival. When telomeres become critically short, cells senesce [@Campisi2003], and the accumulation of these cells has been suggested to result in organismal senescence and death [@Wong2003]. The association between telomere length and senescence has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival or lifespan have been documented in wild populations of several species [@Barrett2013a; @Stier2015]. As yet, however, there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015].

Although the causal role of telomeres in senescence and survival is not yet clear, there is mounting evidence that telomeres can act as biomarkers of individual condition and ageing in wild populations. Specifically, telomeres may be able to provide a measure of the ecological stress that an individual has experienced - a signature that can otherwise be difficult to detect [e.g. @Schultner2014; @Asghar2015; @Bebbington2016]. There is also evidence that telomere length, measured longitudinally in individuals, can increase as well as decrease [@Simons2014; @Bateson2016a], which has important ramifications for our understanding of how telomeres reflect costs. However, such increases in telomere length are often attributed to measurement error [@Steenstrup2013; but see @Bateson2016a], and as such their ecological significance is unknown.

@@

In this study, we test how lifelong telomere dynamics are related to environmental variation across 22 Seychelles warbler cohorts. We first study how telomere length and rates of shortening are related to age and sex across all life stages, and how this relationship varies among cohorts, in order to gain an in-depth understanding of the temporal dynamics of telomere changes. We then examine, within individuals, how telomere length changes with age, and statistically test whether observed increases in telomere length over individual lifespans are larger than can be accounted for by measurement error. Finally, we test how telomere length and shortening are related to a wide range of social and environmental variables in order to gain a fuller understanding of the forces driving telomere dynamics in natural populations.

@@

The Seychelles warbler is a small (~15 g), insectivorous passerine bird with a mean life expectancy of 5.5 years at fledging [@Hammers2013]. The population of *ca* 320 adult birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Komdeur1992; @Richardson2003b; @Spurgin2014a]. This species' main breeding season runs from June-September (though a small proportion of pairs also breed between January–March), when the breeding females on many of the *ca* 110 territories will attempt to breed, laying one or, rarely, two or three eggs [@Komdeur1991]. Breeding attempts are often unsuccessful, and as a result of this low reproductive output, and higher mortality in first-year birds [39% in first-year birds versus 16% in adults; @Brouwer2006], cohort sizes in the Seychelles warbler are typically small (< 50; Table S1). The 22 hatch year cohorts used in this study cover `r min(dd$LayYear)` to `r max(dd$LayYear)` – the time period during which our data and sampling are most complete.

The majority (96%) of individuals are ringed (with an individually numbered metal ring and unique combination of colour rings) within the first year of life, and so are of known age. We aged all birds using information on eye colour at first capture [@Komdeur1991a] and previous capture history [@Richardson2003b]. Within the first year of life, birds are classified as nestlings less than one month old (rounded to one month for analyses), fledglings less than six months old (rounded to six months) or subadults up to one year old (rounded to 10 months). Ages for adult birds were rounded to the nearest year. As Seychelles warblers are non-migratory endemics naturally confined to the island [@Komdeur2004a], an extensive biannual census of birds on Cousin during each breeding season gives accurate measures of local density, social status (e.g. breeder, helper, non-helper) and individual survival [@VandeCrommenacker2011; @Barrett2013a]. Full details of monitoring methods can be found in Brouwer *et al.* [-@Brouwer2012].

@@

For a large subset of birds we had longitudinal data, with two or more samples taken at different ages (n = `r length(unique(c(dd3$BloodID,dd3$BloodID1)))` measurements from `r length(unique(dd3$BirdID))` birds). For these individuals we calculated the within-individual change in RTL by subtracting RTL at time point *t* from RTL at time point *t* + 1 (hereafter $\Delta$RTL, n = `r nrow(dd3)` measurements). Negative values of RTL reflect decreases in telomere length with age, while positive values reflect increases. Individuals were not always caught in the same month, but were generally caught within a 3 month breeding season window.

@@

Using the longitudinal data, we then tested how telomeres change with age in individuals, using LMMs of RTL as a response and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009] as an explanatory variable. We calculated $\Delta$age using log and polynomial transformed age data, and carried out model selection as above, with the exception that we did not model $\Delta$age as a factor (due to a lack of discreet groupings), and mean age was also included in models to partition within-individual *vs* cross-sectional effects [@VandePol2009]. We also excluded the qPCR plate ID random effect from the longitudinal analyses, as each longitudinal measurement was obtained from reactions run on separate plates.

@@

When examining the distribution of longitudinal telomere changes we observed some increases in telomere length with age in individuals. We therefore repeated the qPCR on a large number of samples, using completely separate reactions run on separate plates. We used these repeat measurements to test whether these increases could be explained by measurement error. We calculated the change in RTL between pairs of repeat measurements within the same samples (hereafter $\Delta$RTL~*sample*~; N = `r nrow(subset(ddL,Group == 'Within samples'))` pairs of measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` birds) in exactly the same way as for across samples (hereafter $\Delta$RTL~*individual*~). To test whether greater changes in RTL were observed among individuals compared to among repeat samples, we compared the variance in $\Delta$RTL~*sample*~ and $\Delta$RTL~*individual*~ using a Levene's test. Then, to separately test whether the extent of telomere increases and decreases within individuals were greater than expected by measurement error, we split $\Delta$RTL measurements into groups in which RTL decreased ($\Delta$RTL < 0) and increased $\Delta$RTL ($\Delta$RTL > 0), and tested whether $\Delta$RTL~*individual*~ values were significantly different from $\Delta$RTL~*sample*~ values, using Wilcoxon tests.

We also tested whether consistent telomere lengthening occurred across our dataset using a modified version of the approach developed by Simons *et al.* [-@Simons2014]. Briefly, this approach utilises samples with at least three telomere measurements to compare residual variance in telomere change over time with the overall change in telomere length between the first and last telomere measurements [@Simons2014]. If, in samples that increase in length, the overall increase in telomere length exceeds the residual variance, then telomere lengthening cannot be explained by error [@Simons2014]. If, on the other hand, increases in telomere length are due to measurement error, within-individual residual variance in telomere length is expected to be similar to overall observed increases in telomere length.

@@

For individuals with longitudinal data we repeated the above analyses of telomere dynamics, replacing telomere length with $\Delta$RTL~*individual*~ as the response variable, and including the environmental/social explanatory variables from the first of the two sampling points. We excluded the qPCR plate ID random effect from this analysis (see above), and excluded the cohort ID random effect, as longitudinal telomere dynamics did not differ among cohorts; see results.

##Results

We measured telomere lengths using a total of `r length(unique(dd$BloodID))` unique samples from juvenile and adult Seychelles warblers from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)` (Table S1). Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were `r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))` and `r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))` respectively. Intra-plate repeatability was 0.74 (CI = 0.74,0.75) and 0.73 (CI = 0.71,0.74) for the GAPDH and Telomere Cq values, respectively. Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice at different time points, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`, `r round2(rptall$CI_emp[2])`). Using samples taken from adults greater than one year old, we checked whether RTL was related to sample storage time, and found no evidence of such a relationship (estimate = `r round3(fixef(RTL_ST)['StorageTime'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_ST)['StorageTime',],lessthan = F),collapse = ', ')`).
